---
title: "lab7"
output: pdf_document
date: "2023-02-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Modelling with Stan 

\begin{itemize}
\item Stan is its own language used in r, python etc and need a C++ compiler. 
\item We input our model and it calculates the log posterior density and generates samples from it.
\item It can be confusing to learn at first and you will frequently encounter errors and bugs so don't worry too much.
\end{itemize}

A stan file needs 3 things:
\begin{enumerate}
\item Data: sample data and its size
\item Parameters: What we wish to sample
\item Model: Our likelihood and priors
\end{enumerate}

We need to specify data types such as integers or real numbers and we are also able to impose bounds using "<lower=0>". Its also important to remember to end lines of code with a semi-colon ;. We use the "check" button on our Stan file and want it to return "file.stan is syntactically correct." otherwise we have a problem in our code.

## Example: Fly Wing Length
We wish to generate samples of the length of a species of fly's wings. We are given that 
$$y_i \sim N(\mu, \sigma^2)$$ and we assume the following priors on $\mu$ and $\sigma$ based on a previous study where we find that the average wing length is 1.9mm:
$$\sigma \sim Cauchy(0,1) \textrm{ and } \mu \sim N(\mu_0, \frac{\sigma^2}{\kappa_0})=N(1.9, \frac{\sigma^2}{1})$$. 
We have the following data 
```{r}
y = c(1.64, 1.7, 1.72, 1.74, 1.82, 1.82, 1.9, 2.08)
n = length(y)
```
Now, we want to sample values of $\mu$ and $\sigma^2$:
```{r pressure, echo=FALSE}
# install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
# install_cmdstan()
library(cmdstanr)
library(bayesplot)
```

```{r}
stan_model <- cmdstan_model("example_stan.stan")

stan_fit <-
  stan_model$sample(
    data = list(N = n, y = y, k0=1),
    refresh = 0, show_messages=FALSE)
```
We can inspect our model output in a number of ways. The first is Posterior summary statistics:
```{r}
stan_fit$summary()
```
We will also want to extract draws from our posterior:
```{r}
draws_df <- stan_fit$draws(format = "df")
print(draws_df)
```
We can plot these draws as follows:
```{r}
mcmc_hist(stan_fit$draws("mu"))
mcmc_hist(stan_fit$draws("sigma"))
```








