---
title: "R Notebook"
output: html_notebook
---

#1) Modelling with Stan; Overview 

\begin{itemize}
\item Stan is its own language used in r, python etc and need a C++ compiler. 
\item We input our model and it calculates the log posterior density and generates samples from it.
\item It can be confusing to learn at first and you will frequently encounter errors and bugs so don't worry too much.
\end{itemize}

A stan file needs 3 things:
\begin{enumerate}
\item Data: 1)                2)
\item Parameters: 
\item Model: set distribution of both the 1) likelihood 2) priors 
\end{enumerate}

Notation: 
1) When defining a new paramter/input variable specify the type real/integer 
2) lower bounds on the values: 
3) end each line with: 

#2) Example 1. Fly Wing Length.

Goal: 

Given: 
1) Distribution of $y_i \sim N(\mu,\sigma^2)$
2) Prior on $\mu$: $\sim N(1.9, \frac{\sigma^2}{\kappa_0})$
3) Prior on $\sigma^2$: 

#2a) Create stan file & name it ``example_stan.stan''

#2b) 

We have the next data (obs):
```{r}
y = c(1.64, 1.7, 1.72, 1.74, 1.82, 1.82, 1.9, 2.08)
n = length(y)
```

define model from package 
```{r}
#install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
#install_cmdstan()
library(cmdstanr)
library(bayesplot)

#model 
stan_model <- cmdstan_model("example_stan.stan")

#extract samples 
stan_fit <-
  stan_model$sample(
    data = list(N = n, y = y, k0=1),
    refresh = 0, show_messages=FALSE)
```

Posterior summary 
```{r}
stan_fit$summary()
```

We will also want to extract draws from our posterior:
```{r}
draws_df <- stan_fit$draws(format = "df")
print(draws_df)
```
We can plot these draws as follows:
```{r}
mcmc_hist(stan_fit$draws("mu"))
mcmc_hist(stan_fit$draws("sigma"))
```
q1: does the distribution of mu seem Normal? mean? 



#3) Example 2. Group Exercise: (Similar to Q3 HW)


#1) Given: 
-IQ tests  
-Take sample of n random individuals $y_1, ..., y_n \sim N(\mu,\sigma^2)$. 
-Joint distribution of $\mu$ & $\sigma$ 

-Let $\mu_0 = 100$  (prior) and $\kappa_0 = 1$. Since assume $E[\sigma] = 15$, suppose $a = 15$ and $b=1$. 


#2) Goal: estimate population mean in town.  

##IQ tests observed 
```{r}
y = c(70,85,111,111,115,120,125)
n = length(y)
```


#Step 1): create the appropate stan file and call it IQ_model.stan

#Step 2) extract the samples from stan code 

```{r}
normal_stan_model = cmdstan_model("IQ_model.stan")

stan_fit = normal_stan_model$sample(data = list(#fill in data values 
), refresh = 0, show_messages = FALSE)

samples = stan_fit$draws(format = "df")
```

#Step 3) Make a scatter plot of the posterior distrubtion of the mean mu and the precision 1/sigma^2.

```{r}
mu_samples =
sigma_samples = samples$sigma 
precision_samples =

all_samples = 
```


