---
title: "Lab2"
author: "PSTAT 115, Winter 2022"
date: "Jan 11, 2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
urlcolor: blue
header-includes:
 \usepackage{float}
---

#Step Up.
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo=TRUE, 
                      cache=FALSE, 
                      fig.width=5, 
                      fig.height=5,
                      fig.align='center',
                      fig.pos = 'H')
library(tidyverse)
library(ggplot2)
```

#Useful Peliminaries (Functions) for this Lab. 
```{r}
#1) %>% : 
#1a) :
#2) ggplot(): 
```



#2) Data Generating Process. 

** Main Question: How to model chickweight when given information regarding 1. time 2. chicken id 3. diet? 
  
** Let's begin with the code. 

1. Dataset
```{r}

```

2. Exploring our Data 

2a. Investigating chicken weights. 
```{r}

```


2b. Investigating chicken weights while fixing other covariates. 

Distribution of weight given a fixed time: p(weight|Time = t)
```{r}
df %>% filter(Time == 16) %>% ggplot(aes(weight))+ geom_histogram()
```


Q1: plot chick weights under the next requirements.
  - As a function of time vs weight. (using geom_point)
  - Shade the data points wrt to Diet. 
```{r}
#Answer 

```
  



3. Generate obersavetions (y_1, ..., y_n) from our model. Here, Y_i = Chick Weight (random variable).

** Assumptions: 

  1. Weights are normally distributed.
  
  * 

  2. Effect of Covariates. 
  
  *  
  
  * As diet changes, the rate of increase (alpha) changes.
  
  Final Model:
  
  *Note: diet is categorical (ranges between 1-4). Assume 
  p = (4,.2,.2,.2) (distrbution of each diet category)
```{r}
p = c(4,.2,.2,.2)
for(i in 1:nrow(ChickWeight)){
  Draw diet_i from Categorical(4,p) #sampling a random diet variable accordingly (dis. p) 
  Set alpha_i = alpha_k if diet_i = k, k = 1,2,3,4
  Draw weight_i from Normal(alpha_i * t_i + beta, sigma) 
  
}
```

#3) Baysian Example 

1. Main Question: assume a user searched the word radom. 

2. Equations: See board. 

3. Code:
```{r}
likelihood = c(0.00193, 0.000143, 0.975)
prior = c(7.60e-5, 6.05e-6, 3.12e-7)

unnormalized_posterior = likelihood * prior
posterior = unnormalized_posterior / sum(unnormalized_posterior)
posterior
```




